Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-05-30 manuscript/311 manuscript.tex
Cleaning up

The contributions of this paper are twofold. First, we delve into
the specifics of data curation challenges using the NYC 311 Service
Request Data as a case study. This renowned and frequently viewed 
dataset serves as a prime example for examining key issues in data curation, 
including data validity, consistency, and curation efficiency. 
We illustrate these points with live examples drawn from our processing of the 311 SR data.

@@

\begin{itemize}
 	\item 2003 - NYC 311 system goes live with a phone-based call center only.
   	\item 2009 - 311 online \& mobile apps launched. Today 60\% of all Service Requests come from mobile \& online channels.
   	\item 2019 - Major software upgrade of 311 system with enhanced capabilities, dynamic load-balancing, and cloud-based resilience. 	
      \item 2020 - In August 2020, driven by the COVID pandemic, monthly 311 Service Requests hit an all-time high of 348,463. Seven of the Top 10 busiest days of all time occur in 2020.
      \item 2021 - 311 System is expanded to into the MTA's city subway system, the largest expansion since its inception.
   	\item 2023 - Record high year for 311 with 3.23 million Service Requests.
\end{itemize}

@@

Since the NYC 311 System went live in 2003, the usage has grown at an annual
rate of 4.6\%. For this study, we also looked at a 10-year period (2014-2023) to
observe certain system performance characteristics over a longer time-frame . 
During that time-frame, the number of 311 SRs grew 50\% as shown in the chart below showing yearly SR counts.

@@

This 10-year timeframe (2014-2023) included the COVID pandemic impact. Note that the pandemic
impact reached a peak in the July-September 2020 timeframe, accompanied by additional spikes
in the 2\textsuperscript{nd}half of 2021 as well. The all-time daily high for SRs occurred on Tuesday, 2020-08-04, with a spike of
24,415 SRs, a full 9.8 \textsigma's from the mean.

This below chart shows the 10-year period by calendar month. Note that almost all the months before the COVID pandemic outbreak in Feb/Mar
of 2020 are ``below average'', while most of the months after that period are ``above average'' months. And that trend of higher usage
of the 311 system has continued into the early months of 2024.

@@

A quick look at the corresponding  daily tend over a 5-year period (2019-2023) shows both significant spikes
and significant dips; a noisy graph. While the spikes on 2020-08-04 (24,415 SRs) and 2020-08-05 (19,560 SRs) are certainly COVID related, we were unable to 
ascertain any reason for the significant spike on Saturday, 2023-09-29 when 7,962 SRs were created.

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{Daily.png}
  \caption{Daily SR counts over 5-year period}
  \label{fig:daily-counts}
\end{figure}

Seasonal trends are also prevalent. The months of November-thru-April are ``below average'' months, while the months May-thru-October are ``above average''. 
This seasonal trend is observed throughout the 10-year period.

@@

Of note, the lowest SR counts-per-day occur during the Christmas  holiday period encompassing fully 5 of the Bottom 10 SR counts-per-day. The summer months and early
fall months make up all of the Bottom 10 SR counts-per-day. As noted, the all-time daily high for SRs occurred on Tuesday, 2020-08-04 while the 10-year
low count occurs on Christmas Day in 2016, a Sunday. Note also that the difference between the peak day (24,415) and the minimum day (2965) varies
by a factor of 8X, and well exceeds the average daily count of 7489; all of which creates a challenging  system capacity and performance issue.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{@{}lll|lll@{}}
        \toprule
        \multicolumn{3}{c|}{Top 10 Days} & \multicolumn{3}{c}{Bottom 10 Days} \\
        \cmidrule(r){1-3} \cmidrule(l){4-6}
        Date & Count & Rank & Date & Count & Rank \\
        \midrule
        2020-08-04 & 24415 & 1 & 2014-12-25 & 2965 & 1 \\
        2020-08-05 & 19560 & 2 & 2015-12-25 & 3247 & 2 \\
        2023-09-29 & 17962 & 3 & 2014-04-20 & 3287 & 3 \\
        2020-07-05 & 16916 & 4 & 2015-12-26 & 3405 & 4 \\
        2020-06-21 & 15883 & 5 & 2016-11-06 & 3419 & 5 \\
        2020-06-20 & 15825 & 6 & 2014-07-04 & 3455 & 6 \\
        2020-07-04 & 15794 & 7 & 2014-12-27 & 3534 & 7 \\
        2021-09-02 & 15205 & 8 & 2014-09-14 & 3563 & 8 \\
        2020-06-28 & 14057 & 9 & 2014-02-02 & 3565 & 9 \\
        2021-10-28 & 13575 & 10 & 2016-12-25 & 3587 & 10 \\
        \bottomrule
    \end{tabular}
    \caption{Top 10 and Bottom 10 Days}
    \label{tab:combined_counts}
\end{table}

And although the highest number of SRs occur in July and the lowest in February,
when the \#-of-SRs-per-day metric is computed by adjusting for the differing number of days for each month -- the
busiest month is shown to be June (followed by July and September). Similarly the quietest months are April, March, and 
February. In general, the winter months see fewer SRs while the summer months show 
an increase in activity.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{lS[table-format=7.0]S[table-format=6.0]} % 'l' for left-aligned, 'S' for siunitx number alignment
        \toprule
        \textbf{Month}     & \textbf{Count}   & \textbf{Count per Day} \\ 
        \midrule
        June  & 2454156 & 81805 \\ 
        July  & 2461525 & 79404 \\ 
        September & 2352822 & 78427 \\ 
        October & 2407238 & 77653 \\ 
        August & 2401731 & 77475 \\ 
        May  & 2334466 & 75305 \\ 
        November  & 2236037 & 74535 \\ 
        January   & 2271524 & 73275 \\ 
        December  & 2199765 & 70960 \\ 
        February  & 1996408 & 70795 \\ 
        March & 2174349 & 70140 \\ 
        April  & 2058169 & 68606 \\ 
       \bottomrule
    \end{tabular}
    \caption{Monthly counts ordered by count-per-day}
    \label{tab:monthly_counts}
\end{table}

@@

Two important fields in this analysis effort are the responsible NYC Agency, and the type of complaint.
Given both the size of the City of New York government, it's necessary to identify the responsible Agency as well as the type of complaint in order to 
clarify the responsible party and assist in troubleshooting discrepancies.

When we discovered a data error, we typically observe two trends; either a single (or a few) Agencies are responsible (such as Dept of Transportation), or it is a system-wide issue that follows the general distribution
of  Service Requests (SRs) across the system. Thus we can often identify if it is an Agency specific problem, or a systemic effect affecting all Agencies.

@@

\begin{figure}[H]
  \centering
	  \includegraphics[width=\textwidth]{SRs\_by\_Agency.png}
	  \caption{SR counts by Agency and Cumulative Percentage}

@@

Note the distribution of the complaint\_type; it is heavily skewed to a few select complaints with an exceptionally long tail.
This graph shows the top 20 complaint\_type(s) out of a total of 220.

@@

The top 20 complaints contain several of the eight different types of ``Noise'' complaints (Residential, Commercial, Street, Helicopter, etc.). These Noise-related
complaints total 1.435 million in the 2022-2023 dataset, a full 22\% of all complaint types, which is the most frequently occurring type. (Note: Ilegal Parking is second, followed by Heat/Hot Water.) Noise complaints are also handled by NYPD, as is Illegal Parking and Blocked Driveway complaints, which contributes
significantly to the top ranking of the NYPD as the responsible Agency.

\begin{table}[H]
    \centering
    \footnotesize
    \begin{tabular}{lS[table-format=7]S[table-format=2.2]l} % 'l' for left-aligned, 'S' for siunitx number alignment
        \toprule
        \textbf{complaint\_type} & \textbf{Count} & \textbf{Percentage} & \textbf{Agency} \\ 
        \midrule
        NOISE - RESIDENTIAL        & 675502 & 10.56 & NYPD  \\ 
        NOISE - STREET/SIDEWALK    & 300507 &  4.70 & NYPD  \\ 
        NOISE - COMMERCIAL         & 128892 &  2.02 & NYPD  \\ 
        NOISE - VEHICLE            & 115956 &  1.81 & NYPD  \\ 
        NOISE                      & 100413 &  1.57 & DEP   \\ 
        NOISE - HELICOPTER         &  85345 &  1.33 & EDC   \\ 
        NOISE - PARK               &  16620 &  0.26 & NYPD  \\ 
        NOISE - HOUSE OF WORSHIP   &   2757 &  0.04 & NYPD  \\ 
        \bottomrule
    \end{tabular}
    \caption{Noise-related complaints\_type(s) by count with Agency}
    \label{tab:noisecomplaints}
\end{table}

There are also some  curious and unusual Service Requests in the 311 data: 
tanning, tattooing, trans fat, unsanitary pigeon condition, illegal animal kept as pet, harboring bees/wasps, and the author's favorite
radioactive material. New Yorkers certainly live interesting lives.

The NYC HPD handles about 49\% of what NYPD handles. For  non-NYC readers, 
HPD is the City agency that manages the 177,569 NYC public housing units as well as monitoring
all rental and leased apartment units. HPD collectively serves 528,105 people, a population larger than Atlanta or Miami. 
Hence the large number of Heat/Hot Water and Unsanitary complaints handled by HPD.

@@

\begin{itemize}
	\item Data validation - this effort can span a number of criteria
	\begin{itemize}
		\item Mandatory fields: Certain data fields cannot be empty.
		\item Data-types: Certain fields must be of the correct type, e.g. numeric, character, date, 5 numeric digits, etc. Typically these are identified in a Data Dictionary.
		\item Domain adherence:  Many data fields must adhere to a specific domain of values, e.g. statuses, state names, zip codes, gender. This includes data
		fields that are restricted to certain ranges, such as longitudes and latitudes.
	\end{itemize}   
	\item Structural errors to include naming conventions, extra fields not in the Data Dictionary, or inconsistent data entry.
	For example intermixing blanks, spaces, NA, N/A, and <NA> all to indicate the absence of data.
	\item Redundant, unnecessary, or irrelevant fields
	\item Logical inconsistencies such as related fields that violate the nature of that relationship, e.g. a ``due date'' that is before the ``created date''
	\item Data standardization. Often free-form entry fields can suffer from inconsistent data structures. Efforts to standardize these fields can greatly assist analysis.
	\item Accuracy and precision, which of course are not the same thing 
\end{itemize}

@@

	\subsection{Issues with Zip Codes}
	 Unfortunately some fields proved to be  problematic when comparing the values to a domain of legal values.
	 For example, all zip codes (two fields: zip\_codes and incident\_zip) should be valid as defined by the USPS database
	 which contains 37,946 valid zip codes.

@@

	\subsection{Issues with Police Precincts}
	A curious case also exists when examining the two nearly identical fields - police\_precincts and police\_precinct. Both of those fields are among
	the ``computed'' fields in the dataset. 
	Using \href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}{NYPD Precinct listings} it's possible to determine the valid
	police precincts; they are generally numeric or have numeric representations in the dataset.

@@

The largest errors are shown ``excluding extreme negative values''. We found eight SRs with extremely large negative durations (< -730),
all containing an entry of ``1900-01-01'' as the closed\_date, which generates extremely large negative durations exceeding 44,601 days (122 yrs).
That large of an anomaly can skew statistical results, even though the number of occurrences is small. This is especially curious
as the date ``1900-01-01'' is also the earliest date that Microsoft Excel can accommodate as well as how Excel treats a blank
cell that is designated as a date. All raising suspicions that perhaps these dates were imported into the 311 system from Excel.

@@

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{l l r l}
        \toprule
        \textbf{created\_date} & \textbf{closed\_date} & \textbf{duration} & \textbf{agency} \\
	        \midrule
	        2022-02-11 13:43:05 & 1900-01-01 & -44601.571586 & DHS \\
	        2022-02-11 11:03:38 & 1900-01-01 & -44601.460856 & DHS \\
	        2022-02-11 08:29:49 & 1900-01-01 & -44601.354039 & DHS \\
	        2022-02-11 14:08:08 & 1900-01-01 & -44601.588981 & DHS \\
	        2022-02-11 14:05:16 & 1900-01-01 & -44601.586991 & DHS \\
	        \bottomrule
    \end{tabular}
    \caption{Sample of SRs with extremely large negative durations}
    \label{tab:extreme_negative_durations}
\end{table}

@@

	\subsection{created\_date and closed\_date(s) --  Zero Duration}		
	An even more serious, and more prevalent problem with closed\_date(s) occurs when the closed\_date and created\_date
	are exactly the same -- to the second -- which accordingly creates a \textbf{zero duration'}. Again this is nonsensical, but the
	presence of such zero durations can again severely distort analytics. There are 191,141 such SRs representing 3.1\% of all 
	non-blank data; not an insignificant number. As shown in the chart below, 99\% of the zero duration SRs appear to predominately
	occur in five Agencies (Dept of Mental Health \&Hygine, Dept of Transportation, Dept of Business, Dept of Sanitation,
	and Dept of Environmental Protection.) This is not in-line with the overall Agency distribution of SRs and indicates an
	Agency-specific issue, and the solution likely lies at those Agencies.

@@

	\subsection{created\_date and closed\_date(s) -- Midnight \&Noon}
	One final problem discovered with the created\_date and closed\_date; there appears to be an unusual number of exactly
	midnight and noon date-time values for these two fields. The distribution of SR creation largely follows the clock; lots of
	SRs created during day-light hours and much fewer created in the early hours of the morning. However, there appears
	to be significantly greater numbers of SR both created and closed exactly at midnight, and exactly at noon. The reason this
	issue is worth investigating, goes back to the ``duration'' question -- if SRs are used as a source to measure responsiveness
	for the delivery of services, and if those times are not an accurate reflection of the start/finish of the Service, then the
	duration is in question. Observations during this 2-year period:

	\begin{itemize}
		    \item  there are 7210 SRs created exactly at midnight
		    \item There are 99,779 SRs created exactly at noon.
		    \item There are 105,505 SRs closed exactly at noon. 
		    \item There are 235,347 SRs closed exactly at midnight
	\end{itemize}

	Let's look SRs are created and closed by the hour of the day :

@@

	These ``closed'' charts clearly show significant spikes in the number of SRs at both midnight
	and noon. The midnight spike is clearly out of line with the hours before and after midnight. And
	both the closed-at-noon and closed-at-midnight are approaching 2 standard deviations from the average.

	These unusual patterns of created \& closed SRs on exactly the hours of midnight and noon likely indicate
	some kind of bulk process that perhaps automatically ``closes'' a large number of SRs with a provided
	time-stamp of midnight 00:00:00 or noon 12:00:00. Perhaps there is a similar situation for the ``created'' 
	SRs. So to correct this issue likely requires a close look at what Agencies are providing these suspect SR closed/created times?

	It is not an obvious answer. A visualization of the ``SRs closed at midnight'' shows a distribution that closely mirrors
	the overall assignment of SRs, as the so-called ``Big Six'' Agencies dominate.

	\begin{figure}[H]
		 \centering
		 \includegraphics[scale = 0.65]{closed_at_midnight_chart.png}
		 \caption{SRs closed at Midnight by Agency}
		 \label{fig:midnight-closed}
	\end{figure}

	But the ``SR closed at noon'' visualization shows that just a single Agency, Dept of Sanitation (DSNY),
	is responsible for > 99\% of the SRs closed exactly at noon. This indicates that it is necessary
	to work with DSNY to fully understand what is happening.

	\begin{figure}[H]
		 \centering
		 \includegraphics[scale = 0.65]{closed_at_noon_chart.png}
		 \caption{SRs closed at Noon by Agency}
		 \label{fig:noon-closed}
	\end{figure}

	\subsection{resolution\_action\_update\_date}

\section{Accuracy and precision}\label{sec:precision}
There is one area in particular where the question of precision vs. accuracy immediately arises, and that is with the Latitude and Longitude fields.  Both the 
Longitude and Latitude fields are expresses as a 14-decimal number, e.g. 40.86769186022511 (also the Location field which is a straight concatenation of
latitude and longitude). Given that 1 degree of latitude at the equator is equal to 111.044736 kilometers, the ``1'' at the end of that number represents
approximate 1.1104 nanometer or 1/1,000,000,000 of a meter. A DNA molecule is approximately 2nm in width.

@@

\subsection{Case Study I - Response to Requests for Homeless Person Assistance}

>>>>

The contributions of this paper are twofold. First, we delve into
the specifics of data curation challenges using the NYC 311 Service
Request (SR) Data as a case study. This renowned and frequently viewed 
dataset serves as a prime example for examining key issues in data curation, 
including data validity, consistency, and curation efficiency. 
We illustrate these points with live examples drawn from our processing of the 311 SR data.

@@

\begin{itemize}
 	\item 2003 - NYC 311 system goes live with a phone-based call center only. The ``311'' phone number replaces a myriad of numbers to call for each individual City Agency.
   	\item 2009 - 311 online \& mobile apps launched. Today 60\% of all Service Requests come from mobile \& online channels.
   	\item 2019 - Major software upgrade of 311 system with enhanced capabilities, dynamic load-balancing, and cloud-based resilience. Only major software upgrade since 2004.	
      \item 2020 - In August 2020, driven by the COVID pandemic, monthly 311 Service Requests hit an all-time high of 348,463. Seven of the Top 10 busiest days of all-time occur in 2020.
      \item 2021 - 311 System is expanded to into the MTA's city subway system, the largest expansion since its inception.
   	\item 2023 - Record high year for 311 with 3.23 million Service Requests.
\end{itemize}

@@

Since the NYC 311 System went live in 2003, the usage has grown at an annual
rate of 4.6\%. As part of this study, we also looked at a 10-year period (2014-2023) to
observe system performance over a longer time-frame . During that time-frame, 
the number of 311 SRs grew 50\%.

@@

This 10-year timeframe (2014-2023) includes the COVID pandemic. Note that the pandemic
impact reached a peak in the July-September 2020 timeframe, accompanied by additional spikes
in the 2\textsuperscript{nd}half of 2021 as well. The all-time daily high for SRs occurred on Tuesday, 2020-08-04, 
with a spike of 24,415 SRs, a full 9.8 \textsigma's from the mean.

This chart shows the 2-year period which is the basis for this analysis, by calendar month. 
Note that almost all the months before the COVID pandemic outbreak in Feb/Mar
of 2020 are ``below average'', while most of the months after that period are ``above average''. That level of higher usage
of the 311 system has continued into the early months of 2024.

@@

A quick look at the corresponding  daily tend.  An average day is 8761 SRs.

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{Daily.png}
  \caption{Daily SR counts over 2022-2023 period}
  \label{fig:daily-counts}
\end{figure}

Seasonal trends are also prevalent. The months of November-thru-April are ``below average'' months, while the months May-thru-October are ``above average''. 
This seasonal trend is observed throughout the 10-year period as visualized here.

@@

It is not surprising that the lowest SR counts-per-day occur during the Christmas holiday period. The summer months and early
fall months make up all of the Top 5 SR counts by day. (As a note, the all-time daily high for SRs occurred on Tuesday, 2020-08-04 while the 10-year
low count occurs on Christmas Day in 2016, a Sunday. Note also that the difference between the peak day (24,415) and the minimum day (2965) varies
by a factor of 8X; all of which creates a challenging  system capacity and performance issue.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{@{}lll|lll@{}}
        \toprule
        \multicolumn{3}{c|}{Top 5 Days} & \multicolumn{3}{c}{Bottom 5 Days} \\
        \cmidrule(r){1-3} \cmidrule(l){4-6}
        Date & Count & Rank & Date & Count & Rank \\
        \midrule
        2020-08-04 & 24,415 & 1 & 2014-12-25 & 2965 & 1 \\
        2020-08-05 & 19,560 & 2 & 2015-12-25 & 3247 & 2 \\
        2023-09-29 & 17,962 & 3 & 2014-04-20 & 3287 & 3 \\
        2020-07-05 & 16,916 & 4 & 2015-12-26 & 3405 & 4 \\
        2020-06-21 & 15,883 & 5 & 2016-11-06 & 3419 & 5 \\
        \bottomrule
    \end{tabular}
    \caption{Top 5 and Bottom 5 Days}
    \label{tab:combined_counts}
\end{table}

@@

Two important fields in this analysis effort are the responsible NYC Agency, and the type of complaint.
Given the size of the City of New York government, it is necessary to identify the responsible Agency as well as the type of complaint in order to 
identify the responsible party and assist in troubleshooting discrepancies.

When we discovered a data error, we typically observe one of two trends; either a single (or a few) Agencies are responsible (such as Dept of Transportation), 
or it is a system-wide issue that follows the general distribution of  SRs across the system. 
Thus we can often identify if it is an Agency specific problem, or a systemic issue.

@@

\begin{figure}[H]
  \centering
	  \includegraphics[width=\textwidth]{SRs_by_Agency.png}
	  \caption{SR counts by Agency and Cumulative Percentage}

@@

The distribution of complaint\_type skews to a few select complaints accompanied by a long tail.
This graph shows the top 20 complaint\_type(s) comprising 70\% out of a total of 220.

@@

The top 20 complaints contain several of the eight different types of ``Noise'' complaints (Residential, Commercial, Street, Helicopter, etc.). These Noise-related
complaints total 1.435 million in the 2022-2023 dataset, a full 22\% of all complaint types, which, when consolidated, is the most frequently occurring complaint\_type.
Ilegal Parking is second, followed by Heat/Hot Water. Noise complaints are handled by NYPD, as is Illegal Parking and Blocked Driveway complaints, which contributes
significantly to the top ranking of the NYPD as the lead responsible Agency, receiving a full 43\% of all SRs. 
Housing, Preservation, \& Development are 2\textsuperscript{nd}, with 21\%

For  non-NYC residents, HPD is the City agency that manages the 177,569 NYC public housing units as well as monitoring
all City rental and leased apartment units. HPD collectively serves 528,105 people, a population larger than Atlanta or Miami. 
Hence the large number of Heat/Hot Water and Unsanitary complaints handled by HPD.

\begin{table}[H]
    \centering
    \footnotesize
    \begin{tabular}{lS[table-format=7]S[table-format=2.2]l} % 'l' for left-aligned, 'S' for siunitx number alignment
        \toprule
        \textbf{complaint\_type} & \textbf{Count} & \textbf{Percentage} & \textbf{Agency} \\ 
        \midrule
        NOISE - RESIDENTIAL        & 675,502 & 10.56 & NYPD  \\ 
        NOISE - STREET/SIDEWALK    & 300,507 &  4.70 & NYPD  \\ 
        NOISE - COMMERCIAL         & 128,892 &  2.02 & NYPD  \\ 
        NOISE - VEHICLE            & 115,956 &  1.81 & NYPD  \\ 
        NOISE                      & 100,413 &  1.57 & DEP   \\ 
        NOISE - HELICOPTER         &  85,345 &  1.33 & EDC   \\ 
        NOISE - PARK               &  16,620 &  0.26 & NYPD  \\ 
        NOISE - HOUSE OF WORSHIP   &   2757 &  0.04 & NYPD  \\ 
        \bottomrule
    \end{tabular}
    \caption{Noise-related complaints\_type(s) by count with Agency}
    \label{tab:noisecomplaints}
\end{table}

There are also some curious and unusual Service Requests in the 311 data: 
tanning, tattooing, trans fat, unsanitary pigeon condition, illegal animal kept as pet, harboring bees/wasps, and the authors' favorite
radioactive material. New Yorkers certainly live interesting lives.

@@

\begin{itemize}
	\item Data validation - this effort can span a number of criteria
	\begin{itemize}
		\item Mandatory fields: Certain data fields cannot be empty.
		\item Data-types: Certain fields must be of the correct type, e.g. numeric, character, date, 5 numeric digits, etc. Typically these are identified in a Data Dictionary.
		\item Domain adherence:  Many data fields must adhere to a specific domain of values, e.g. statuses, state names, zip codes, gender. This includes data
		fields that are restricted to certain ranges, such as longitudes and latitudes.
	\end{itemize}   
	\item Structural errors to include naming conventions, extra fields not in the Data Dictionary, or inconsistent data entry.
	For example intermixing blanks, spaces, NA, N/A, and \textless{}NA\textgreater{} all to indicate the absence of data.
	\item Redundant, unnecessary, or irrelevant fields
	\item Logical inconsistencies such as related fields that violate the nature of that relationship, e.g. a ``due date'' that is before the ``created date''
	\item Data standardization. Often free-form entry fields can suffer from inconsistent data structures. Efforts to standardize these fields can greatly assist analysis.
	\item Accuracy and precision, which of course are not the same thing 
\end{itemize}

@@

	\subsection{Issues with Zip Codes}
	\label{sec:zip-codes}
	 Unfortunately some fields proved to be  problematic when comparing the values to a domain of legal values.
	 For example, all zip codes (two fields: zip\_codes and incident\_zip) should be valid as defined by the USPS database
	 which contains 37,946 valid zip codes.

@@

	\subsection{Issues with Police Precincts}
	\label{sec:police-precincts}
	A curious case also exists when examining the two nearly identical fields - police\_precincts and police\_precinct. Both of those fields are among
	the ``computed'' fields in the dataset. 
	Using \href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}{NYPD Precinct listings} it's possible to determine the valid
	police precincts; they are generally numeric or have numeric representations in the dataset.

@@

The largest errors are shown ``excluding extreme negative values''. We found eight SRs with extremely large negative durations (\textless{} -730),
all containing an entry of ``1900-01-01'' as the closed\_date, which generates extremely large negative durations exceeding 44,601 days (122 yrs).
That large of an anomaly can skew statistical results, even though the number of occurrences is small. This is especially curious
as the date ``1900-01-01'' is also the earliest date that Microsoft Excel can accommodate as well as how Excel treats a blank
cell that is designated as a date. All raising suspicions that perhaps these dates were imported into the 311 system from Excel.

@@

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{l l r l}
        \toprule
        \textbf{created\_date} & \textbf{closed\_date} & \textbf{duration} & \textbf{agency} \\
	        \midrule
	        2022-02-11 13:43:05 & 1900-01-01 & -44602 & DHS \\
	        2022-02-11 11:03:38 & 1900-01-01 & -44602 & DHS \\
	        2022-02-11 08:29:49 & 1900-01-01 & -44601 & DHS \\
	        2022-02-11 14:08:08 & 1900-01-01 & -44602 & DHS \\
	        2022-02-11 14:05:16 & 1900-01-01 & -44602 & DHS \\
	        \bottomrule
    \end{tabular}
    \caption{Sample of SRs with extremely large negative durations}
    \label{tab:extreme_negative_durations}
\end{table}

@@

	\subsection{created\_date and closed\_date(s) --  Zero Duration}		
	An even more serious, and more prevalent problem with closed\_date(s) occurs when the closed\_date and created\_date
	are exactly the same -- to the second -- which accordingly creates a \textbf{zero duration'}. Again this is nonsensical, but the
	presence of such zero durations can again severely distort analytics. There are 191,141 such SRs representing 3.1\% of all 
	non-blank data; not an insignificant number. As shown in the chart below, 99\% of the zero duration SRs appear to predominately
	occur in five Agencies (Dept of Mental Health \&Hygiene, Dept of Transportation, Dept of Business, Dept of Sanitation,
	and Dept of Environmental Protection.) This is not in-line with the overall Agency distribution of SRs and indicates an
	Agency-specific issue, and the solution likely lies at those Agencies.

@@

	\subsection{created\_date and closed\_date(s) -- Midnight \&Noon}
	An additional  problem discovered with the created\_date and closed\_date fields; there appears to be an unusually large number of 
	SRs created or closed at exactly midnight (00:00:00) and noon (12:00:00), to the second. The distribution of SR creation and closure 
	largely follows the work-day clock with lots of SRs created during day-light hours, and fewer SRs 
	created in the early hours of the morning. However, here there appears to be significantly greater numbers of SR closed exactly at midnight and exactly at noon,
	as well as a significant number of SRs created exactly at noon. The reason for this anomaly is worth investigating 
	as it affects the SR ``duration'' question; if SR durations are used to measure responsiveness
	for the delivery of services, and if those times are not an accurate reflection of the start/finish of the Service, then the use of
	duration as a service quality metric is called into question. Observations during this 2-year period include:

	\begin{itemize}
		    \item There were 235,347 SRs closed exactly at midnight
		    \item There were 105,505 SRs closed exactly at noon. 
		    \item There were 99,779 SRs created exactly at noon 
	\end{itemize}

	Here is a visualization of SRs closed and created by the hour of the day :

@@

	The closed\_date chart show significant spikes in the number of SRs at both midnight
	and noon with significant counts at exactly those times, to the second. The midnight spike is certainly suspect compared to the hours just before and after midnight where
	there is normally a pronounced slump in SR closures. And both the closed-at-noon and closed-at-midnight are 
	approaching 2\textsigma's from the hourly average. The created\_date shows a maximum value at noon also with a significant count occurring exactly at noon, again to 
	the second.

	These unusual patterns of created \& closed SRs at exactly the hours of midnight and noon perhaps indicates
	the presence of a bulk process; one that perhaps automatically ``closes'' (or ``creates'') a large number of SRs with a provided
	time-stamp of midnight (00:00:00) or noon (12:00:00). If so, then addressing this issue likely requires a closer look at those 
	Agencies providing these suspect SR closed/created times. Clearly the desirable state is that the created\_date and closed\_date fields
	reflect the true opening and closing of the SRs, such that an accurate ``duration'' can be computed which accurately
	represents the SR life-cycle.

	Looking at the distribution by Agency for the ``closed-exactly-at-midnight'' shows that \textgreater{}90\% of these suspect SRs come from just two Agencies - Dept. of 
	Buildings (DOB) and Dept. of Sanitation (DSNY). If you include HPD and DOT, that accounts for 99\% of the suspect SR closures.

	\begin{figure}[H]
		 \centering
		 \includegraphics[scale = 0.65]{closed_at_midnight_chart.png}
		 \caption{SRs Closed Exactly at Midnight by Agency}
		 \label{fig:midnight-closed}
	\end{figure}

	Hoever, the ``SR closed at noon'' visualization shows that  a single Agency, DSNY,
	is responsible for \textgreater{}99\% of the SRs ``closed-exactly-at-noon''. Accordingly, any effort to resolve this issue
	necessitates working with DSNY to fully understand what is occurring.

	\begin{figure}[H]
		 \centering
		 \includegraphics[scale = 0.65]{closed_at_noon_chart.png}
		 \caption{SRs Closed Exactly at Noon by Agency}
		 \label{fig:noon-closed}
	\end{figure}

	The 99,779 SRs created-exactly-at-noon also appears to originate with the Dept of Sanitation (DSNY) which
	is responsible for \textgreater{}99\% of the suspect SRs.

	\begin{figure}[H]
		 \centering
		 \includegraphics[scale = 0.65]{created_at_noon_chart.png}
		 \caption{SRs Created Exactly at Noon by Agency}
		 \label{fig:noon-created}
	\end{figure}

	It is difficult to determine if these observations are indeed anomalies. They could be the result of normal, expected
	311 SR activity. But at least by observation that appears unlikely.

	\subsection{resolution\_action\_update\_date}
	When an SR is updated, by any means on any field, the software automatically populates the resolution\_action\_update\_date. Analyzing that field
	it is apparent that some of the SR updates are happening long, long after the SR is closed. In this dataset there are eight SRs that have 
	resolution\_action\_update\_date values of 44,601 days after the closed\_date. The problem lies not with the resolution\_action\_update\_date, but
	rather with the closed\_date values for these 8 SRs, all of which are in the format of ``1900-01-01''. It has previously been discussed that these
	inaccurate dates may possibly be coming from an Excel spreadsheet as that is the minimum date which Excel can handle. Here is a sample
	of those SRs:

\begin{table}[H]
	    \centering
	    \footnotesize
	    \begin{tabular}{cccc}
	        \hline
	        Agency & Closed Date & Resolution Action Updated Date & Post Closed Update Duration (days) \\
	        \hline
	        DHS & 1900-01-01 & 2022-02-12 00:58:43 & 44602 \\
	        DHS & 1900-01-01 & 2022-02-11 14:28:59 & 44601 \\
	        DHS & 1900-01-01 & 2022-02-11 14:17:41 & 44601 \\
	        DHS & 1900-01-01 & 2022-02-11 13:45:59 & 44601 \\
	        DHS & 1900-01-01 & 2022-02-11 13:22:12 & 44601 \\
	        \hline
	    \end{tabular}
	    \caption{Extremely long Resolution Updates}
	    \label{tab:resolution-updates}
	\end{table}

	These obvious inaccurate date values are removed from the analysis. But even so, there are some very late
	updates to SRs, with some updates occurring almost two years after the SR is closed.   There are a total of 7460 SRs that are 
	updated \textgreater{}30 days and \textless{}730 days after the closed\_date. The median of these late (\textless{}30 days)
	post-closed resolution\_action\_update\_date(s) is 84 days. However, the mean resolution\_action\_update\_date 
	for all of the SRs is only 0.39 days (9.3 hours) after the SR is closed. It is not known if this is normal behavior
	or an area that may require further investigation.

	\begin{table}[H]
    \centering
    \footnotesize
	    \begin{tabular}{cccc}
	        \hline
	        Agency & Closed Date & Resolution Action Updated Date & Post Closed Update Duration (days) \\
	        \hline
	        TLC & 2023-05-25 14:21:12 & 2023-10-19 10:57:02 & 147 \\
	        TLC & 2023-07-12 11:03:46 & 2023-12-07 09:00:05 & 148 \\
	        TLC & 2023-08-03 10:34:23 & 2024-01-29 10:00:37 & 179 \\
	        TLC & 2022-04-04 10:54:39 & 2022-06-02 12:14:11 & 59 \\
	        TLC & 2022-12-05 15:23:45 & 2023-02-02 09:06:45 & 59 \\
	        \hline
	    \end{tabular}
    \caption{Long post-closed Resolution Updates}
    \label{tab:resolution}
	\end{table}

	Here is a look at the spread of the post-closed resolution\_action\_update\_date values those SRs 
	that are \textgreater{}30 days and \textless{}730 days. There are some significant outliers

	\begin{figure}[H]
		 \centering
		 \includegraphics[scale = 0.65]{post_closed_violin.png}
		 \caption{Post-Closed resolution\_action\_update\_date >30 days }
		 \label{fig:resolution-violin}
	\end{figure}

	If we look at a breakout of these (late?) update dates by Agency, we can see that the vast majority li with two
	Agencies: Taxi \& Limo Commission (TLC) and Dept of Sanitation (DSNY) which account for 96\% of the SRs in
	question.  Is there a valid reason that TLC and DSNY would be updating an SR over month (or months) after
	the SR has been closed? Perhaps, but it seems to merit closer investigation.

	\begin{figure}[H]
		 \centering
		 \includegraphics[scale = 0.65]{postClosedBarChart.png}
		 \caption{Post-Closed resolution\_action\_update\_date >30 days by Agency }
		 \label{fig:resolution-by-agency}
	\end{figure}

\section{Accuracy and precision}\label{sec:precision}
There is one area in particular where the question of precision vs. accuracy immediately arises, and that is with the Latitude and Longitude fields.  Both the 
Longitude and Latitude fields are expresses as a 14-decimal number, e.g. 40.86769186022511 (also the Location field which is a straight concatenation of
latitude and longitude). Given that 1 degree of latitude at the equator is equal to 111.044736 kilometers, the ``1'' at the end of that number represents
approximate 1.1104 nanometer or 1/1,000,000,000 of a meter. For reference a DNA molecule is approximately 2nm in width.

@@

\section{Redundant \& Duplicate fields}\label{sec:duplicates}
During this analysis, several redundant fields were observed.

\subsection{latitude \& longitude and the location fields}  The location field is a pure concatenation of the latitude and longitude fields, 
with a comma an parenthesis added. Example:

	\begin{itemize}
		\item  latitude: 40.768456429488
		\item  longitude: -73.9575661888774
		\item  location: (40.768456429488, -73.95756618887745)
	\end{itemize}

The inclusion of the location field seems questionable as the data is arguably more difficult to extract than the two specific fields, especially for
software programs.
